\naslov{Matchings}

\begin{definition}
  A vertex set $S \subseteq V$ is an \pojem{independent set} of the graph $G =
  (V, E)$ if the induced subgraph $G[S]$ is empty.
  The maximum cardinality of an independent set is the \pojem{independence
	number} $\alpha(G)$.
\end{definition}

\begin{definition}
  A vertex set $T \subseteq V$ is a \pojem{vertex cover} if every edge has at least one
  of its endings in $T$.
  The maximum cardinality of a vertex cover is the \pojem{vertex cover number}
  $\beta(G)$.
\end{definition}

\begin{definition}
  An edge set $M \subseteq E$ is a \pojem{matching} if for every distinct $e_1,
  e_2 \in M$, edges $e_1$ and $e_2$ have no common ending.
  The maximum cardinality of a matching is the \pojem{matching number}
  $\alpha'(G)$.
\end{definition}

\begin{definition}
  An edge set $C \subseteq E$ is an \pojem{edge cover} if every vertex of $G$ is
  covered by at least one edge from $C$.
  If the minimum degree $\delta(G)$ is at least $1$, we can define the
  \pojem{edge cover number} $\beta'(G)$ as the minimum cardinality of an edge
  cover.
\end{definition}

\vprasanje{Define the independence number, the vertex and edge cover number, and
  the matching number.}

\begin{remark}
  The complement of an independent set is a vertex cover, so $\alpha(G) +
  \beta(G) = n(G)$ in every graph $G$.
  In a maximum matching, every edge must be covered by different vertices, so
  $\alpha'(G) \le \beta(G)$.
  We can similarly argue that $\alpha'(G) \le \nicefrac{n(G)}{2} \le \beta'(G)$
  and $\alpha(G) \le \beta'(G)$.
\end{remark}

\begin{theorem}[Gallai]
  If $\delta(G) \ge 1$, then $\alpha'(G) + \beta'(G) = n(G)$.
\end{theorem}

\begin{proof}
  Take a maximum matching $M$ in $G$ and let $V(M)$ be the vertices covered by
  $M$.
  For every vertex not covered by $M$, we can take an incident edge and add it
  to $M$.
  This gives an edge cover with
  \[
	\abs{M} + \abs{\comp{V(M)}} = \abs{M} + (n - 2 \abs{M}) = n - \abs{M}
  \]
  edges.
  Since $\abs{M} = \alpha'(G)$, this implies $\beta'(G) + \alpha'(G) \le n(G)$.

  Now take a minimum edge cover $C$.
  We claim that for every edge in $C$, at least one end is covered only once by
  $C$.
  For suppose that $uv \in C$ is an edge and both $u$ and $v$ are covered by
  other edges in $C$.
  If we remove $uv$, then $C \setminus \{uv\}$ is a smaller cover, which is a
  contradiction.

  The induced subgraph $G[C]$ is then a forest of stars.
  Suppose it consists of $k$ components.
  We get $\abs{C} = n-k$, since in a tree, the number of vertices is $1$ more
  than the number of edges.
  A matching is obtained by choosing one edge from every star, which gives
  $\alpha'(G) + \beta'(G) \ge n(g)$, thus completing the proof.
\end{proof}

\vprasanje{State and prove Gallai's theorem.}

\begin{definition}
  Let $M$ be a matching.
  A path $v_1 v_2 \ldots v_k$ is an \pojem{$M$-alternating path} if
  the edges along the path alternate between $M$ and $\comp{M}$.
  An $M$-alternating path is \pojem{$M$-augmenting} if neither end of the path
  is covered by $M$.
\end{definition}

\begin{remark}
  Such a path cannot start or end with an edge from $M$, and the endpoints
  cannot be part of an edge in $M$.
\end{remark}

\begin{proposition}
  If $G$ is a graph, $M$ is a matching and there exists an $M$-augmenting path
  $P$, then $M$ is not a maximum matching.
\end{proposition}

\begin{proof}
  Suppose $P = v_1 \ldots v_k$.
  We know the first and last edge are not in $M$, so $\abs{E(P) \cap \comp{M}} =
  \abs{E(P) \cap M} + 1$.
  Now let $M' = M \oplus E(P)$ be the symmetric difference of $M$ and $E(P)$.
  This is clearly a matching.
  We know that $\abs{M'} = \abs{M} + 1$, so $M$ cannot be maximum.
\end{proof}

\vprasanje{How can you construct a larger matching from an augmenting path?}

\begin{definition}
  A \pojem{König-Egerváry} graph is a graph $G$ with $\alpha'(G) = \beta(G)$.
\end{definition}

\begin{theorem}[König]
  Let $G$ be a bipartite graph.
  Then $\alpha'(G) = \beta(G)$.
  Additionally, if $M$ is a matching in $G$ and there is no $M$-augmenting
  path, $M$ is a maximum matching.
\end{theorem}

\begin{proof}
  Let the partite classes of $G$ be $A$ and $B$.
  Suppose that $M$ is a matching for which there is no $M$-augmenting path in
  $G$.
  Define $X$ as the set of all vertices in $A$ that are not covered by the
  matching, and $Y$ the vertices in $B$ not covered by $M$.
  Additionally, let $B_1$ be the set of vertices in $B$ that can be reached by
  an $M$-alternating path from $X$, and similarly, let $A_1$ be the set of
  vertices of $A$ which can be reached from $X$ by an $M$-alternating path.
  Finally, define $B_2 = B \setminus (B_1 \cup Y)$ and $A_2 = A \setminus (A_1
  \cup X)$.

  Observe that on an $M$-alternating path from $X$, any edge from $A$ to $B$ is
  in $\comp{M}$ and any edge from $B$ to $A$ is in $M$.
  Out matching provides a one-to-one mapping between $A_1$ and $B_1$ and between
  $A_2$ and $B_2$, so $\abs{A_1} = \abs{B_1}$ and $\abs{A_2} = \abs{B_2}$.
  We also know that $\abs{A_1} + \abs{A_2} = \abs{M}$.
  Now consider the possible edges between the defined vertex sets.

  There is no edge between $X$ and $Y$, since that would be a (trivial)
  $M$-augmenting path.
  There are also no edges between $X$ and $B_2$, since an edge $xb$ is an
  $M$-alternating path, implying $b \in B_1$.
  So the only edges from $X$ lead to $B_1$.

  There are also no edges between $A_1$ and $Y$, because we can construct an
  $M$-augmenting path with such an edge.
  If $a \in A_1$, then there is an alternating path from $X$ to $a$, which we
  could extend with a $a$-to-$Y$ edge to get an augmenting path.
  Finally, there are no edges between $A_1$ and $B_2$, since that would give an
  alternating path from $X$ to the $B_2$-vertex as before.

  Then $T = B_1 \cup A_2$ is a vertex cover with
  $\abs{T} = \abs{B_1} + \abs{A_2} = \abs{A_1} + \abs{A_2} = \abs{M}$.
  We have thus constructed a vertex cover with $\abs{M}$ vertices, giving
  \[
	\beta(G) \le \abs{T} = \abs{M} \le \alpha'(G).
  \]
  The other inequality holds in the general case, so this completes the proof.
\end{proof}

\vprasanje{State and prove König's theorem.}

\begin{corollary}
  If $G$ is a bipartite graph, then $\alpha(G) = \beta'(G)$.
\end{corollary}

\begin{definition}
  Let $G$ be a bipartite graph with partite classes $A$ and $B$.
  \pojem{Hall's condition} holds for the set $A$ if for every $S \subseteq A$,
  \[
	\abs{S} \le \abs{N(S)} = \abs{ \bigcup_{u \in S} N(u) }.
  \]
\end{definition}

\begin{theorem}[Hall]
  If $G$ is a bipartite graph with partite classes $A$ and $B$, then there
  exists a matching that covers $A$ if and only if Hall's condition holds for
  $A$.
\end{theorem}

\begin{proof}
  Let $M$ be a matching covering $A$ and $S \subseteq A$.
  We can take the pairs matched by $M$
  \[
	B_S = \{ v \in B \such \text{$v$ is covered by an edge in $M$} \}.
  \]
  Clearly, $\abs{S} = \abs{B_S}$ and $B_S \subseteq N(S)$, so $\abs{S} \le
  \abs{N(S)}$.

  For the other implication, suppose there is no matching covering $A$.
  Divide the sets $A$ and $B$ as in the proof of König's theorem, using some
  matching $M$.
  Since the matching doesn't cover $A$, $X$ is not empty.
  Now consider $S = A_1 \cup X$.
  All edges from $S$ lead into $B_1$, so $N(S) = B_1$, but $\abs{S} = \abs{A_1}
  + \abs{X} > \abs{B_1} = \abs{N(S)}$.
\end{proof}

\vprasanje{State and prove Hall's theorem.}

\begin{definition}
  A matching $M$ is a \pojem{perfect matching} if it covers all vertices.
\end{definition}

\begin{corollary}
  In a bipartite graph $G$, there is a perfect matching if and only if $\abs{A}
  = \abs{B}$ and $A$ satisfies Hall's condition.
\end{corollary}

\begin{definition}
  Let $G$ be a bipartite graph with partite classes $A,B$, and $S \subseteq A$.
  The \pojem{deficiency} of $S$ is $\operatorname{def}(S) = \abs{S} -
  \abs{N(S)}$.
\end{definition}

\begin{theorem}
  Let $G$ be a bipartite graph with partite classes $A$ and $B$.
  If $M$ is a maximum matching in $G$, it covers
  \[
	\alpha'(G) = \abs{A} - \max_{S \subseteq A} \operatorname{def}(S)
  \]
  vertices of $A$.
\end{theorem}

\begin{theorem}
  If $G$ is a regular bipartite graph, then $G$ has a perfect matching.
\end{theorem}

\begin{proof}
  The number of edges in the graph is $k \cdot \abs{A} = k \cdot \abs{B}$, so
  $\abs{A} = \abs{B}$.
  Let $S \subseteq A$.
  The number of edges between $S$ and $N(S)$ is exactly $k \cdot \abs{S}$.
  Every neighbour $u \in N(S)$ has exactly $k$ neighbours, at most $k$ are in
  $S$.
  So at most $k \cdot \abs{N(S)}$ edges are between $S$ and $N(S)$, implying
  $\abs{S} \le \abs{N(S)}$.
\end{proof}

\vprasanje{Show that a regular bipartite graph has a perfect matching.}

\begin{theorem}
  Let $M$ be a matching in $G$.
  Then there is an $M$-augmenting path in $G$ if and only if $M$ is not a
  maximum matching in $G$.
\end{theorem}

\begin{proof}
  We've already proved the right implication.
  Suppose there is a matching $M'$ with $\abs{M'} > \abs{M}$.
  Consider the symmetric difference $M \triangle M'$ and denote $G' = G[M
  \triangle M']$.
  Clearly the maximum degree $\Delta(G') \le 2$, from which we know that the
  components of $G'$ are all paths or cycles.

  Any cycle must alternate between edges from $M$ and $M'$, so it is an even
  cycle, and contains the same number of edges from the two matchings.
  In any path of even length, there must be the same number of edges in $M$ and
  in $M'$.
  And finally, in a path of odd length, one of the two sets has an extra edge
  compared to the other.
  Since $\abs{M'} > \abs{M}$, there must be a path with more edges in $M'$ than
  in $M$.
  Label it $G_1'$.

  This component is an $M$-augmenting path in $G$, since if either of its
  endpoints are covered by $M$, they must also be covered by the same edge in
  $M'$, but then $M'$ wouldn't be a matching.
\end{proof}

We can find the maximum matching in polynomial time, with so-called
\enquote{Blossom algorithms}, which find an augmenting path in $O(m \sqrt{n})$.
This also means we can determine the edge-cover number $\beta'(G)$ in polynomial
time.

\podnaslov{Tutte's theorem}

\begin{definition}
  A component of a graph $G$ is \pojem{odd} if it has an odd number of vertices.
  We denote the number of odd components in $G$ with $o(G)$.
\end{definition}

\begin{theorem}[Tutte]
  A graph $G$ has a perfect matching if and only if for any $S \subseteq V(G)$,
  $\abs{S} \ge o(G - S)$ holds.
  This is called Tutte's condition.
\end{theorem}

\begin{proof}
  Left to right: Let $S \subseteq V(G)$ and $M$ be a perfect matching in $G$.
  Let $H_1, \ldots, H_k$ be the components of $G - S$.
  If $H_i$ is an odd component, then there exists at least one $M$-edge between
  $V(H_i)$ and $S$.
  Therefore $\abs{S} \ge o(G-S)$.

  Right to left:
  Suppose that Tutte's condition holds for a graph $F$ but there is no perfect
  matching in $F$.
  Now add edges to $F$ so that this still holds after the addition, and let $G$
  be a maximal such graph on $n(F)$ vertices.
  If we consider Tutte's condition on $S = \varnothing$, we see that there are
  no odd components in $G$, which means $n(G)$ is even.

  Now take any edge in the complement of $G$.
  Since $G$ is maximal, $G+e$ is not a counterexample, so it either has a
  perfect matching, or Tutte's condition does not hold for it.
  We know that for any $S \subseteq V(G)$, $\abs{S} \ge o(G-S)$.
  After having added an edge, at most one pair of components in $G-S$ is joined.
  If this was a pair of odd components, $o(G-S)$ has reduced, and in all other
  cases, it has remained the same.
  This means that $G+e$ must satisfy Tutte's condition, so it must have a
  perfect matching as it is not a counterexample.

  We will consider two cases.
  Let $U$ be the set of universal vertices in $G$, that is, the vertices
  adjacent to every other vertex, and let $H_1, \ldots, H_k$ be the components
  of $G - U$.
  In the first case, suppose every $H_i$ induces a complete graph.
  We will construct a perfect matching.
  For the even components, we may create a matching within the component, and
  for the odd components, we may connect the one remaining vertex with $U$.
  We are left over with an even number of vertices in $U$ (since $n$ is even),
  which we may form a matching with, as $G[U]$ is a complete graph.
  So in this case, $G$ is not a counterexample, which is a contradiction.

  Now suppose there is a non-complete component $H_i$.
  Then there exist vertices $x, y, z$ for which $xy \notin E(H_1)$ but both $xz$
  and $yz$ are in $E(H_1)$.
  There is also another vertex $w \in V(G)$ for which $zw \notin E(G)$, since
  $z$ is not a universal vertex.
  Let $G_1 = G + xy$ and $G_2 = G + zw$.
  We know both these graphs have perfect matchings, which must include $xy$ and
  $zw$ respectively.
  Denote the perfect matchings with $M_1$ and $M_2$ and consider $G[M_1
  \triangle M_2]$.
  Note that every non-isolated vertex in this graph is of degree $2$, since it
  is covered by both perfect matchings.
  These vertices must of course appear in even cycles.

  If $xy$ and $zw$ belong to different components, then we may choose edges in
  each component, and the edges deleted by the symmetric difference, to form a
  perfect matching, and we can avoid taking $xy$ or $zw$.
  This is a contradiction.
  Alternatively, if $xy$ and $zw$ belong to the same component, they appear in
  the same cycle.
  Without loss of generality they appear in the order $zwxy$ (but not
  necessarily adjacent).
  To form a new perfect matching, we will take the edge $xz$, and one edge set
  from each side of the cycle.
  This avoids both new edges $xy$ and $zw$, so we have a contradiction.
\end{proof}

We also have the Berge-Tutte formula, which states that a maximum matching in
$G$ leaves uncovered exactly
\[
  \max_{S \subseteq V(G)} \{o(G - S) - \abs{S}\}
\]
vertices.

\podnaslov{Factors}

\begin{definition}
  A \pojem{factor} is a spanning subgraph (it contains all vertices).
  A \pojem{$k$-factor} is a $k$-regular spanning subgraph.
\end{definition}

\begin{definition}
  A \pojem{cubic graph} is a $3$-regular graph.
\end{definition}

\begin{theorem}[Petersen]
  Every bridgeless cubic graph has a $1$-factor.
\end{theorem}

\begin{proof}
  We will prove that Tutte's condition holds.
  Take $S \subseteq V(G)$.
  Since the graph is $3$-regular, the number of edges between $S$ and $\comp{S}$
  is $\abs{E(S,\comp{S})} \le 3 \abs{S}$.
  If $H_i$ is an odd component in $G - S$, then $E(V(H_i), S)$ contains at least
  one edge.
  Note that
  \[
	2 m(H_i) + \abs{E(H_i, S)} = 3 n(H_i) = \sum_{v \in H_i} \deg_G(v)
  \]
  where $m(H_i)$ is the number of edges in $H_i$.
  Since the RHS is odd, $\abs{E(H_i, S)}$ must be odd.
  Also, $\abs{E(H_i, S)} \ne 1$, since that would be a cut edge (a bridge)
  otherwise.
  Therefore, $\abs{E(H_i, S)} \ge 3$.
  Then $3 \abs{S} \ge \abs{E(S, \comp{S})} \ge 3 \sigma(G - S)$.
\end{proof}

We can also improve this theorem, for example allowing for precisely one bridge
in the graph, and the statement still holds.
Or we could require that all cut edges lie on the same path, and it would still
hold.

\begin{theorem}
  If $G$ is a $k$-regular graph and $k$ is even, then $G$ has a $2$-factor.
\end{theorem}

\begin{proof}
  We can limit ourselves to connected graphs.
  By Euler's theorem, there exists an Eulerian circuit $C$ in $G$.
  We can fix a direction for this $C$ and orient the edges according to it.
  From this, we obtain a directed graph $\vec{G}$.
  For every vertex $v$, $C$ enters and exits $v$ exactly $k/2$ times.

  Define a bipartite graph $F_G$ with partite classes $A = \{a_1, \ldots, a_n\}$
  and $B = \{b_1, \ldots, b_n\}$ for $n = n(G)$ and the edge set
  \[
	a_i b_j \in E(F_G) \iff v_i v_j \in E(\vec{G}).
  \]
  Clearly, $F_G$ is a $\frac{k}{2}$-regular graph, so it has a perfect matching
  $M$.
  Define a spanning subgraph in $G$ with the following edge set $Q$:
  \[
	v_i v_j \in Q \iff a_i b_j \in M \lor a_j b_i \in M.
  \]
  Every vertex $v_i$ is covered exactly twice by $Q$.
\end{proof}

% LocalWords:  Egerváry König König's Tutte's Tutte bridgeless Eulerian
